---
title: "Working with R and RStudio"
author: "Jason Hilton"
institute: "University of Southampton"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "custom.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: center, middle,inverse

# RStudio


```{r setup, message=F, warning=F, echo=F}
library(ggplot2)
library(magrittr)
library(dplyr)
library(knitr)
library(tidyr)
library(DiagrammeR)
library(ggthemes)
library(svglite)

opts_chunk$set(fig.align="center",
               fig.retina=2,message=F, warning=F,
               fig_caption=F, echo=F)

theme_set(theme_grey(base_size=18))
```


---
# What is RStudio for? 

R is a powerful standalone programming language for statistical analysis.

However, the built-in Graphical User Interface (GUI) is rather minimal.

- RStudio provides a powerful Integrated Development Environment for R
- Syntax highlighting and checking ('linting')
- Project organisation facilities
- Code completion
- Integration with a range of other tools
  + Rmarkdown for integrated code and text documents
  + Terminal window
  + Facilities for python, javascript and stan files
  + Debugging and profiling tools
  + Integration with different graphical devices

--

Some may prefer a good text editor (Sublime, Geany) and the terminal,
or another IDE.

- Emacs - through the Emacs Speaks Statistics (ESS) package 
- Visual studio


--
Jupyter notebooks are also R-friendly!

---
class: center, middle,inverse

# Project Organisation


---
# Principles

- Projects should be reproducible from start to finish
- Projects should be documented

---
# Warning!



> If the first line of your R script is

> `setwd("C:\Users\jenny\path\that\only\I\have")`

> I will come into your office and SET YOUR COMPUTER ON FIRE 🔥.

From [Jenny Bryan's blog post.](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/)

--

Not everyone agrees; this **is** a common idiom in R.

Projects organised in this way are **not easily reproducible**.
- 'Future me' won't be able to run this code if I have changed machine
- Colleagues working on other machines will not be able to run this code
- Confusing if different scripts are run from different paths

---
# Project organisation

What is a sensible way of organising projects?

- Create a top level folder for each project
- All paths are written relative to this top level

### Not this:

```{R, eval=F, echo=T}

read.csv("C:/obscure/path/to/some/random/project/data/file.csv")

```

### This

```{R, eval=F,echo=T}
read.csv("./data/file.csv")
```


---
# Rstudio projects

Rstudio helps with this structure through projects.

- Create a new project in your project folder


- When this project is open, Rstudio will automatically start R in this folder


- Some settings can also be provided on a per-project basis


--

These tools help maintain a more reproducible workflow

- You should be able to run the scripts that carry out your analysis from the command line from the top-level project folder


```{bash, eval =F}
cd path/to/my project
Rscript scripts/create_viz.R
```

- *If you have a `setwd` command in your project, this will not work on other machines*

---
# Folder structure for simple projects


.pull-left[
<iframe src="Project_folder.html" height=1000 width=300 style="border:none;"></iframe>
]


.pull-right[

- Top-level project folder

<div style="line-height:130%;">
    <br>
</div>

- The `R` folder will contain user-written functions for sourcing

<div style="line-height:90%;">
    <br>
</div>

- The `scripts` folder will contain the files which carry out the body of the work.

<div style="line-height:10%;">
    <br>
</div>

- Data stored locally can be saved in the `data` folder

<div style="line-height:10%;">
    <br>
</div>

- Results, including plots, are outputed to the `results` folder

]



---
class: center, middle, inverse
# Workflow


---
# Another Warning!

> If the first line of your R script is

> `rm(list = ls())`

> I will come into your office and SET YOUR COMPUTER ON FIRE 🔥.

From [Jenny Bryan's blog post.](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/)

---
# Maintaining a clean environment

The `rm(list=ls())` expression will remove all objects you have created from the global environment.


--
But anything added to your R environment by loading packages will remain!

- Your analysis may not be reproducible



--
- **Restart R** when you restart an analysis
- All libraries should be loaded at the top of your script
- Do not save and load `.Rdata` when opening R


--
Environment managers can help with making sure package versions remain stable within projects


- The [renv](https://rstudio.github.io/renv/articles/renv.html) package is one such manager



---
# Workflow outputs

Save outputs at different stages of your analysis
- Each script should create one output
- The subsequent steps use outputs from previous steps


.pull-left[
- Save specific R objects using `saveRDS` rather than `save`
- Load and assign data using `readRDS` rather than `load`


- This makes it clear in a script where each data object has come from!
- It also prevents overwriting existing `R` objects
]
.pull-right[
```{r, eval=F,echo=T}
saveRDS(an_object, 
        file="file_name.rds")

```

```{r,eval=F,echo=T}
loaded_object <- readRDS(
  file="object.rds")
```
]


--
This allows you to build up an analysis pipeline, enabling each step to be treated in isolation.

---
class: middle, inverse
# Naming conventions


>    There are only two hard things in Computer Science: cache invalidation and  naming things.
>   -- Phil Karlton

[Via Martin Fowler](https://martinfowler.com/bliki/TwoHardThings.html)


---
# Use meaningful names

Understanding code is much easier if functions and variables have meaningful names.

### Obscure names
```{r, echo=T,eval=F}
anls(svdf)
```


--
### Meaningful names
```{r, echo=T, eval=F}
calculate_hazard(survival_df)
```


--
I would recommend doing this even if it is only you who will be using the code

- It's good to get into good habits
- You never know when code might be needed in other contexts



---
# Conventions

- Try not to overwrite built-in variables or functions e.g. `mean` or `fit`
- Use verbs to describe functions
- Don't use dots in variable or function names
 + Some legacy code has this notation
 + However, it clashes with syntax for object-orientated programming in R
- Spaces around operators (`+`, `-`, `/`, and so on), and after commas
- Indent inside functions and `if` statements


More information [here](http://adv-r.had.co.nz/Style.html)






